Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-31 manu-draft.qmd
more edits from RN, mostly for sensitivity analysis

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for 1 day, $n \approx 52$ for 7 days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate. The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1. Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other parameters. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period. About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown. For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a 1-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high. The lines indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot). Similarly, median values were much more variable for the 1 day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges. Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.

The actual model results for the best and worst performing combinations of prior distributions for the 1 day, 7 day, and 30 day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series. Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters. In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter. As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. Gas exchange (D) was similar between the input time series, suggesting that lack of precision in the $b$ parameter had a minimal influence regardless of noise.

@@

![A subset of the synthetic time series and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters. The original synthetic time series and the same time series with instrument and tidal noise is shown in (a) and the estimated instrument and tidal noise is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}

>>>>

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each of five model outputs ($D$, $DO$, $R$, $a$, and $P$), each of 16 unique combinations of priors, and each of 3 optimization periods. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for comparability among the optimization periods. The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1. Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). This result makes sense, given that the $b$ parameter was fixed at the actual value used to create the synthetic time series. Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`).

By optimization period, median NSE values across all model outputs increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all model outputs were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period. About half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the same combination was observed for the 30 day optimization period with the exception that $\sigma$ was high for $R$. Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that model withs sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic results. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown. For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a 1-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high. The lines indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Dissolved oxygen was generally modeled with high skill for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot). Similarly, median values were much more variable for the 1 day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges. Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.

The actual model results for the best and worst performing combinations of prior distributions for the three optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all model outputs in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the model outputs were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b), (d), and (f). In such cases, the $DO$ simulation (fourth row) can still be quite good as a result of compensating errors in $a$ and $R$. The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series. Further, the worst model for the 30 day optimization period in subplot (f) had the same priors as the 7 day optimization period, but was able to better reproduce the synthetic results, suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were dervied from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the residual and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors justified in @sec-ebasemod (equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with noise (residual and tidal) to those from the original synthetic time series without noise. Only results for $a$ and $R$ are shown Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters. In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter. As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. Gas exchange (D) was similar between the input time series, suggesting that lack of precision in the $b$ parameter had a minimal influence regardless of noise.

@@

![A subset of the synthetic time series and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters. The original synthetic time series and the same time series with residual and tidal noise is shown in (a) and the estimated residual and tidal noise added to the synthetic time series is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}
